---
title: "ECON 464: Parental Education and Gender Differences in Career Choices: An Analysis of Legal Professions among US College Graduate"
output:
  html_document: default
  pdf_document: default
date: "2024-05-06"
author: Chi Vu, Dilnoza Khuseynova
---

The research aims to answer the question of whether the education background of an individual’s parents influence the choice of career path, especially in law?


```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyverse)
library(tidyr)
library(stargazer)
library(httr) # Import from web

```

```{r}
#load full NSCG 2021 dataset
# Data: NSCG
url <- "https://wadelitt.com/wp-content/uploads/2024/02/nscg21.csv"
response <- httr::GET(url, timeout(300))  # Sets timeout to 300 seconds

if (status_code(response) == 200) {
  # Read directly from the content of the response
  data_content <- content(response, "text")
  df_nscg21 <- read.csv(text = data_content)
} else {
  print("Failed to download the file")
}

nscg21 <- df_nscg21

# Clean Data --------------------------------------------------------------

# baind   | Indicator of whether respondent received BA degree
# baind = "N" | Did not receive BA (n=997)
nscg21 <- nscg21 %>% filter(baind != "N")

# lwnvr   | Never Worked
# lwnvr = "Y"  | (n=908)
# Respondent checked box "Never worked for pay or profit" 
nscg21 <- nscg21 %>% filter(lwnvr != "Y")

#filter for lawyers and judges
df_law <- nscg21 %>% filter(n3ocpr == "781200" | n3oclst == "781200" ) # creating main dataframe, with lawyers and judges professionals
```

1480 out of 104392 participants are or have been lawyers and judges.

## Data Exploration - df_law

```{r degree distribution}
summary(df_law$dgrdg)

# Convert education level columns to factors with appropriate labels
df_law$dgrdg <- factor(df_law$dgrdg,
                       levels = 1:8,
                       labels = c("Less than high school",
                                  "High school diploma",
                                  "Some college/vocational",
                                  "Bachelor's degree",
                                  "Master's degree",
                                  "Professional degree",
                                  "Doctorate",
                                  "Not applicable"))


# Create a bar chart to show the distribution of highest degree types
ggplot(df_law, aes(x = dgrdg, fill = dgrdg)) +
  geom_bar(color = "black") +
  scale_fill_brewer(palette = "Set3", name = "Degree") +
  labs(title = "Distribution of Highest Degree Types Among Lawyers and Judges",
       x = "Highest Degree Obtained", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r age distribution}

summary(df_law$age)

ggplot(df_law, aes(x = agegr, fill = gender)) +
  geom_bar(position = position_dodge(width = 2.5), stat = "count", color = "black") +
  scale_fill_manual(values = c("M" = "lightblue", "F" = "salmon")) +
  labs(title = "Age Distribution of Lawyers and Judges by Gender", x = "Age Group", y = "Frequency") +
  scale_x_discrete(limits = df_law$agegr, labels = function(x) paste0(x, "-", x+4)) +
  theme_minimal()


```

```{r gender proportion}
# Calculate percentages for each gender
df_law_percent <- df_law %>%
  count(gender) %>%
  mutate(percentage = n / sum(n) * 100)

# Create pie chart with percentages
ggplot(df_law_percent, aes(x = "", y = percentage, fill = factor(gender))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") +
  labs(title = "Gender Distribution Among Lawyers and Judges", fill = "Gender") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), position = position_stack(vjust = 0.5)) +
  theme_void()

```

```{r parents education}
# Any value outside this range or NA can be set to 8 ("Not applicable") as a default

df_law$eddad <- ifelse(is.na(df_law$eddad) | !df_law$eddad %in% 1:8, 8, df_law$eddad)
df_law$edmom <- ifelse(is.na(df_law$edmom) | !df_law$edmom %in% 1:8, 8, df_law$edmom)

# Convert education level columns to factors with appropriate labels
df_law$eddad <- factor(df_law$eddad,
                       levels = 1:8,
                       labels = c("Less than high school",
                                  "High school diploma",
                                  "Some college/vocational",
                                  "Bachelor's degree",
                                  "Master's degree",
                                  "Professional degree",
                                  "Doctorate",
                                  "Not applicable"))

df_law$edmom <- factor(df_law$edmom,
                       levels = 1:8,
                       labels = c("Less than high school",
                                  "High school diploma",
                                  "Some college/vocational",
                                  "Bachelor's degree",
                                  "Master's degree",
                                  "Professional degree",
                                  "Doctorate",
                                  "Not applicable"))

# Reshape data to long format and adjust parent_type
df_law_long <- df_law %>%
  pivot_longer(
    cols = c("eddad", "edmom"),
    names_to = "parent_type",
    values_to = "education_level"
  ) %>%
  mutate(parent_type = recode(parent_type, 
                              "eddad" = "Father", 
                              "edmom" = "Mother"))


# Create a side by side bar chart
ggplot(df_law_long, aes(x = education_level, fill = parent_type)) +
  geom_bar(position = position_dodge(width = 0.9), color = "black") +
  scale_fill_manual(values = c("Father" = "skyblue", "Mother" = "salmon"), name = "Parent") +
  labs(title = ("Parents' Education Level Distribution Among Lawyers and Judges"),
       x = "Education Level", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```



## Logistic Regression
```{r, include=FALSE}
# Specify a CRAN mirror
options(repos = c(CRAN = "https://cloud.r-project.org"))

# Load the necessary package
install.packages("glm2")
library(MASS)  # For logistic regression
library(car)   # For VIF function - multicolinearity check
```


```{r Convert necessary categorical variables to factors, include=FALSE}
# Creating binary variable: 1 if lawyer/judge, 0 otherwise 
nscg21$lawyer <- ifelse(nscg21$n3ocpr == "781200" | nscg21$n3oclst == "781200", 1, 0)

# Convert necessary categorical variables to factors
nscg21 <- nscg21 %>%
  mutate(
     eddad = factor(eddad, levels = 1:8,
                           labels = c("Less than high school",
                                  "High school diploma",
                                  "Some college/vocational",
                                  "Bachelor's degree",
                                  "Master's degree",
                                  "Professional degree",
                                  "Doctorate",
                                  "Not applicable")),
    edmom = factor(edmom,  levels = 1:8,
                           labels = c("Less than high school",
                                  "High school diploma",
                                  "Some college/vocational",
                                  "Bachelor's degree",
                                  "Master's degree",
                                  "Professional degree",
                                  "Doctorate",
                                  "Not applicable")),
    gender = factor(gender),
    spowk = factor(spowk, levels = c(1, 2, 3, "L"), 
                   labels = c("Yes, full-time",
                                     "Yes, part-time",
                                     "No",
                                     "Skip")),
    racethm = factor(racethm, levels = 1:7,
                     labels = c("Asian", 
                                "American Indian/Alaska Native", 
                                "Black", 
                                "Hispanic", 
                                "White", 
                                "Native Hawaiian/Other Pacific Islander", 
                                "Multiple Race"))
         )
```

```{r logit with all possible variables}
# Fit the basic logistic regression model with all possible variables
logit_mod1 <- glm(lawyer ~ eddad + edmom + gender + spowk + chun12 + racethm + age + bthus + facadv + facben + facchal + facind + facloc + facresp + facsal + facsec + facsoc, 
              data = nscg21, 
              family = binomial(link = "logit"))

# Summary of the model
summary(logit_mod1)

```

Coefficient Significance:

- Parental Education (`eddad`, `edmom`): Both coefficients are statistically significant, suggesting a positive association between higher parental education levels and the likelihood of becoming a lawyer. This aligns with our hypothesis that emphasize the role of parental education in career outcomes.

- Demographic Factors (`gender`, `racethm`, `bthus`): Gender is not significant, indicating it does not distinguish between lawyers and non-lawyers in the dataset. 
-> Gender is variable of focus, making it a control variable

- The significant effect of being born in the U.S. (`bthusY`) and certain racial categories (e.g., Hispanic) may highlight demographic influences.

- Age: Age is significant, implying that career decisions to become a lawyer may correlate with age, possibly reflecting career development stages or generational effects.
Job Factors:

- Several job satisfaction factors (`facadv`, `facsal`, etc.) were included, but many show non-significance or lead to singularities. We are focusing on variables with siginificance in the refined model: `facben`, `facresp`, and `facsoc`.

Refining the model:

```{r converting facxxx variables, include=FALSE}
nscg21$facben <- factor(nscg21$facben, levels = c(1, 2, 3, 4, "L"), labels = c("Very important",
                                                                                "Somewhat important",
                                                                                "Somewhat unimportant",
                                                                                "Not important at all",
                                                                                "Logical Skip"))

nscg21$facresp <- factor(nscg21$facresp, levels = c(1, 2, 3, 4, "L"), labels = c("Very important",
                                                                                "Somewhat important",
                                                                                "Somewhat unimportant",
                                                                                "Not important at all",
                                                                                "Logical Skip"))

nscg21$facsoc <- factor(nscg21$facsoc, levels = c(1, 2, 3, 4, "L"), labels = c("Very important",
                                                                                "Somewhat important",
                                                                                "Somewhat unimportant",
                                                                                "Not important at all",
                                                                                "Logical Skip"))
```

```{r refined logit model}
# refined model
logit_mod2 <- glm(lawyer ~ eddad + edmom + gender + age + racethm + bthus + facben + facresp + facsoc,
                  family = binomial(link = "logit"),
                  data = nscg21)
summary(logit_mod2)
```

This model refinement from the more complex initial model has simplified the predictors while maintaining significant insights into what influences the probability of becoming a lawyer.

- Higher educational levels of parents, particularly a professional degree, significantly increase the likelihood of the child being a lawyer. 

- Age and being U.S. born also significantly influence the likelihood, with age increasing the odds and U.S. born status showing a strong positive effect. 

- Some workplace factors such as the importance of benefits and social aspects are significant predictors, whereas the importance of responsibility shows a negative effect. 



```{r}
# Check for multicolinearity
vif_vals <- vif(logit_mod2)
print(vif_vals)
```


Looking at the VIF values, none of the variables exceed the common thresholds of concern for multicollinearity (typically 5 or 10 for VIF, ). However, the variables `eddad` and `edmom` have slightly higher VIF values than others (around 2.8-2.9), although these aren't typically alarming levels for logistic regression.


```{r male logit, include=FALSE}
# Filter data for males
data_male <- subset(nscg21, gender == "M")
logit_male <- glm(lawyer ~ eddad + edmom + age + racethm + bthus + facben +  facresp + facsoc,
                  family = binomial(link = "logit"),
                  data = data_male)
summary(logit_male)
```

```{r female logit, include=FALSE}
# Filter data for females
data_female <- subset(nscg21, gender == "F")
logit_female <- glm(lawyer ~ eddad + edmom + age + racethm + bthus + facben +  facresp + facsoc,
                    family = binomial(link = "logit"),
                    data = data_female)
summary(logit_female)
```


### Model Comparisons

Detailed report in submitted paper.

```{r table: gender comparison}
# Create a comparison table - gender
stargazer(logit_male, logit_female, 
          title = "Comparison of Logistic Regression Models between Gender",
          type = "text",  # Change to "html" or "latex" for different output formats
          model.names = TRUE,  # Include names of models
          out = "gender_comparison_table.html")  # Saves output to an HTML file, remove for console output

```

```{r overall vs gender}
# Comparing full sample with genders
stargazer(logit_mod2, logit_male, logit_female, 
          type="text",
           title = "Comparison of Logistic Regression Models between Gender")
```

